Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-13 doc/thesis/discussion.tex
Intro summary; finished ABC

\begin{figure}[ht]
  \centering
  \includegraphics{derivatives.pdf}
  \caption[
    First and second derivatives of epidemic growth curves at time of sampling
    for various values of $I$ and $N$.
  ]{
    First and second derivatives of epidemic growth curves at time of sampling
    for various values of \gls{I} and \gls{N}. Networks were simulated under
    the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and \gls{N} varied 
    along the values shown on the $x$-axis. Transmission trees were sampled at
    the time when \gls{I} nodes were infected ($y$-axis). Logistic growth
    curves were fit to epidemic trajectories derived from the transmission
    trees, and their first and second derivatives were calculated at the time
    of sampling. Contours show first derivatives, while colours indicate second
    derivatives. Values of \gls{I} and \gls{N} used in simulation experiments
    with kernel-assisted \gls{ABC} are indicated by diamonds.
  }
  \label{fig:derivs}
\end{figure}

@@

For all datasets we examined, the posterior mean estimates for \gls{alpha} were
sub-linear, ranging from 
    0.27
to 
    0.73.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. They found \gls{alpha} values
ranging from 0.26 to 0.62, depending on the gender and time period considered. 
Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
this range under both priors, as were the estimates for the
\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.

Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
populations whose primary risk factor for \gls{HIV} infection was heterosexual
contact. \citeauthor{de2007preferential} explicitly excluded reported
homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
heterosexual contact is the primary mode of transmission in Botswana where the
study was done. In the first of the two papers describing the Botswana
study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
was gender-biased, being composed of approximately 75\% women. Our estimate of
\gls{alpha} for these data was X or X, depending on the prior on \gls{m};
\citeauthor{de2007preferential} estimated 0.54, 0.57, and 0.29 for 3-year,
5-year, and lifetime partnership networks respectively for the female portion
of their sample.

For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
populations had a higher estimated \acrlong{PA} power than the other datasets
(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
in Brooklyn, USA, collected between 1991 and
1993~\autocite{friedman2006social}. They found that the the \gls{IDU} network 
resembled a \gls{BA} network much more closely than other social and sexual 
networks, and offered sociological explanations for the apparent \acrlong{PA}
dynamics in this population. Importantly, from a public health perspective,
the authors asserted that the removal of \emph{random} individuals from
\gls{IDU} networks may have the paradoxical effect of increasing the network's
epidemic susceptibility. When low-degree nodes are removed, as would occur
during a police crackdown, their network neighbours may turn to well-known
community members for advice or supplies, thus increasing the connectivity of
these high-degree nodes.

%For both priors, the estimated prevalence was extremely high, in fact higher
%than the estimated HIV prevalence in the sampled region. The authors indicated
%that the source of the samples was a town in close proximity to the country's
%capital city, and suggested that there may have been a high degree of migration
%and partner interchange between the two locations. It is possible that the
%contact network underlying the subtree we investigated includes a much larger
%group based in the capital city, which would explain the high estimate of $I$.
%There is no clear explanation for the discrepancy between the two priors for
%the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
%cluster and therefore was a good candidate for the BA model. However, nearly
%all the posterior density was assigned to $m = 1$ when this value was allowed,
%indicating that the network was more likely to have an acyclic tree structure.

{\color{blue}\uline{
The estimates of the prevalence \gls{I} were largely robust to the choice of
prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
\textcite{niculescu2015recent} \uline{reported that 494 new HIV infections were
diagnosed among the studied population (\glspl{IDU} in Bucharest) during the
study period. Our estimate of the prevalence was higher (X), although the 95\%
\gls{HPD} interval did contain the value 494. In addition, it is not
unreasonable that a substantial fraction of the \gls{HIV}-positive individuals
in an \gls{IDU} population could be undiagnosed, given that these populations
are often marginalized.}

\uline{The individuals sampled by} \textcite{wang2015targeting} \uline{were
recruited by following a prospective cohort of 2000 \gls{MSM} in Beijing,
China. Of the 2000, 179 new infections were identified during the study period.
This number is much lower than the estimated prevalence of X obtained with
\software{netabc}, however the authors did not claim to have sampled the entire
sexual network. In a nationwide survey,} \textcite{wu2013hiv} \uline{estimated
that there were 24,198 \gls{MSM} living in Beijing, of whom 5.7\%, or 1379,
were \gls{HIV}-infected, which is within the 95\% confidence interval of our
estimate of \gls{I}.}

\textcite{li2015hiv} \uline{studied an \gls{MSM} population in Shanghai, China.
Despite the apparent similarity to the context of the
}\textcite{wang2015targeting} \uline{study, the estimated \gls{HIV} prevalence
of the } \citeauthor{li2015hiv} \uline{data was higher (X). The aforementioned
survey~\autocite{wu2013hiv} estimated that there were 14,511 \gls{MSM} living
in Shanghai, with an \gls{HIV} prevalence of 6.8\% or approximately 987
individuals. This was within the 95\% confidence interval of the estimate
obtained with \software{netabc}. It is worth noting that} \citeauthor{li2015hiv}
\uline{claimed that there were 80,000 \gls{MSM} living in Shanghai, which is a
significantly different estimate than that obtained by} \textcite{wu2013hiv}.

\textcite{cuevas2009hiv} \uline{reported that 620 newly infected and 1500
chronically infected patients had been sampled during the study period, for a
total of 2120. This is substantially greater than the point estimate of X we
obtained with \gls{ABC}, even without considering the likely possibility that
there were more infected than sampled individuals. Again, this error may in
part be due to model misspecification, which highlights the importance of
selecting a model appropriate to the dataset at hand.} }

Our use of the \gls{BA} model makes several simplifying assumptions. First, we
assume homogeneity across the network with respect to node behaviour and
transmission risk. In reality, the attraction to high-degree nodes seems likely
to vary among individuals, as does their risk of transmitting or contracting
the virus. We have also assumed that all transmission risks are symmetric,
which is clearly false for all known modes of \gls{HIV} transmission, and that
infected individuals never recover but remain infectious indefinitely. These
assumptions were made for the purpose of keeping the model as simple as
possible, since this is the very first attempt to fit a contact network model
in a phylodynamic context. However, the Gillespie simulation algorithm built
into \software{netabc} can handle arbitrary transmission and removal rates
which need not be homogeneous across the network. Moreover, it is possible to
use \gls{ABC} to fit a model which relaxes some or all of these assumptions,
which may be a fruitful avenue for future investigation.  Despite the possible
misspecification, our estimates of the power law exponent $\gamma$ were within
the range of values reported in the literature (\cref{tab:gamma}).

>>>>

\begin{figure}[ht]
  \centering
  %\includegraphics{derivatives.pdf}
  \caption[
    First and second derivatives of epidemic growth curves at time of sampling
    for various values of $I$ and $N$.
  ]{
    First and second derivatives of epidemic growth curves at time of sampling
    for various values of \gls{I} and \gls{N}. Networks were simulated under
    the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and \gls{N} varied 
    along the values shown on the $x$-axis. Transmission trees were sampled at
    the time when \gls{I} nodes were infected ($y$-axis). Logistic growth
    curves were fit to epidemic trajectories derived from the transmission
    trees, and their first and second derivatives were calculated at the time
    of sampling. Contours show first derivatives, while colours indicate second
    derivatives. Values of \gls{I} and \gls{N} used in simulation experiments
    with kernel-assisted \gls{ABC} are indicated by diamonds.
  }
  \label{fig:derivs}
\end{figure}

@@

### 2016-06-13 doc/thesis/intro.tex
Intro summary; finished ABC

As discussed in \cref{subsec:smcsamp}, the
choices of the kernels $K$ and $L$ is problem-specific, and so appropriate
kernels must be chosen for \gls{ABC}. Several options have been
proposed~\autocite{beaumont2009adaptive, sisson2007sequential,
del2012adaptive}.

\begin{algorithm}
  \caption{\gls{ABC}-\gls{SMC}.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\Pr(\theta) \Pr(z \mid \theta)$
      \Comment{Draw $\theta$ from the prior and simulate $z$ from $\theta$}
      \State $w^{(k)} \gets 1/n$
    \EndFor
    \For {$i = 2$ to $i_{\max}$}
      \Comment{$i_{\max}$ is a user-specified number of iterations}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
        \Comment Extend the $k$th particle
        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
      \If {$\ESS(w) < T$}
        \Comment{$T$ is a user-defined threshold}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
    \EndFor
  \end{algorithmic}
  \label{alg:abcsmc}
\end{algorithm}

@@

{\color{blue}\uline{ Our method integrates the four distinct research areas
  just described: phylogenetics, contact networks, \acrlong{SMC}, and
  \acrlong{ABC}. The first two topics together form the problem domain.
  Phylogenetic data is the input to our method, while estimates of the
  parameters of contact network models are the desired output. The latter two
  topics define the algorithm and statistical framework that our inference
method will use. }}

>>>>

As discussed in \cref{subsec:smcsamp}, the choices of the kernels $K$ and $L$
is problem-specific, and so appropriate kernels must be chosen for \gls{ABC}.
Several options have been proposed~\autocite{beaumont2009adaptive,
sisson2007sequential, del2012adaptive}. {\color{blue}\uline{With an
appropriately chosen kernel, the weight update} (\cref{eq:smcwt}) \uline{will
simplify into a computable expression.} \textcite{sisson2007sequential} and
\textcite{beaumont2009adaptive} \uline{both suggest random walk kernels for
$K$, where each particle is
perturbed according to a Gaussian distribution. The backwards kernels $L$ are
chosen to approximately minimize the variance in importance weights. In this
thesis, we use an \gls{MCMC} kernel, as proposed by} \textcite{del2012adaptive}.
\uline{The associated backwards kernels and weight updates are discussed in the
next chapter} (\cref{subsec:adaptsmc}). \uline{With generic kernels $K$ and
$L$, the \gls{ABC}-\gls{SMC} algorithm is almost identical to the \gls{SMC}
sampler} (\cref{alg:smcsamp}), \uline{with $\gamma_i$ replaced with $\pi_i$.}
Therefore we will not repeat the full algorithm here.}

@@

\glsreset{ABC}
\glsreset{SMC}
\glsreset{BA}

{\color{blue}\uline{In} \cref{sec:obj}, \uline{we outlined three research
objectives that will be addressed in this thesis. First, we aim to develop a
method for fitting contact network models to estimated transmission trees.
Although transmission trees can be estimated for any epidemic by thorough
contact tracing, viral diseases are the most useful context for this method.
Transmission and sequence evolution occur on similar time scales for RNA
viruses, resulting in viral phylogenies whose shapes are heavily constrained by
the transmission process. The study of this interaction between evolution and
epidemiology is called \emph{phylodynamics}. Phylodynamic methods make it
possible to estimate transmission trees from viral sequence data; these
estimated trees form the input data for our method.}

\uline{From these estimated transmission trees, our method's purpose is to
estimate the parameters of a contact network model. Rather than assuming a
homogeneously mixed population, as most epidemiological models do, network
models take the more realistic view that human populations are structured. That
is, contacts which allow for transmission may occur in a nonrandom way, rather
than between every pair of individuals in the population. A network model
parameterizes this structure. In particular, our second research objective is
to characterize our ability to fit the \gls{BA} model, which incorporates
preferential attachment to generate networks with realistic degree
distributions.}

\uline{Our method will apply \gls{ABC}, a simulation-based approach, for
fitting these models. Simulating a transmission tree according to a network
model is straightforward: a network can be generated according to the model,
and the spread of an epidemic can be simulated over the network and recorded in
a transmission tree. \gls{ABC} uses the concordance between these simulated
transmission trees and the ``true'' estimated tree as an indicator of parameter
credibility. The closer the simulated transmission trees appear to the true
tree, the more weight is assigned to the associated network parameters.}

\uline{\gls{ABC} can be implemented by at least three algorithms, but the one
we choose to apply in this work is \gls{SMC}. \Gls{SMC} uses a population of
parameter ``particles'' to approximate a distribution of interest, in this case
the approximate posterior distribution targeted by \gls{ABC}. After running the
\gls{ABC}-\gls{SMC} algorithm, statistics on the model parameters can be
approximated by the weighted population of particles. For example, a weighted
average would give an approximate expected value for each parameter.}

\uline{Thus, our method integrates four research topics: phylogenetics, contact
networks, \acrlong{SMC}, and \acrlong{ABC}. The first two topics together form
the problem domain. Phylogenetic data is the input to our method, while
estimates of the parameters of contact network models are the desired output.
The latter two topics define the algorithm and statistical framework that our
inference method will use. }}
